---
title: "BIOS 731 HW 1 Commute Analysis"
format: 
  html:
    embed-resources: true
    toc: true
    toc-depth: 2
    toc-title: Contents
    toc-location: left
    df-print: kable
    theme: sandstone
    highlight: haddock
    code-fold: true
    code-overflow: wrap
    knitr:
      opts_chunk:
        echo: true,
        include: true
        message: false,
        warning: false
        results: markup
        fig.align: center
        fig.height: 5
        fig.width: 7
        fig.path: "../results/"
        R.options:
          scipen: 999
          knitr.kable.NA: ""
editor: source
---

```{r}
#| label: setup
#| include: false

# Clear memory
rm(list = ls()); gc()

library(tidyverse)
library(magrittr)
library(here)
library(knitr)
library(gtools)
library(broom)
library(corrplot)
```

# Overview

This document runs a reproducible analysis for BIOS 731 Homework 1. We download data from the American Community Survey S0201 table at [data.census.gov](https://data.census.gov/table/ACSSPP1Y2024.S0201?t=Commuting:Education&g=010XX00US$0500000&moe=false&tp=true&tableFilters=ag-Grid-AutoColumn~(Margin+of+Errorundefined)). This table provides county-level demographics for a selection of counties across the United States, including information such as total population, total population with a disability, median household income, and average commute time to work. 

We are interested in the demographics that may be associated with average commute time, since there are likely a wide variety of factors that may govern this part of county residents' lives. To explore this, we select a subset of fields from this table and run regressions to identify meaningful associations.

# Load and Clean Data

First, we download data from the American Community Survey. After navigating to the S0201 table, we download the 2024 data as a zip file. This provides one csv file (with the suffix "-Data") containing the data for each county, one csv file (with the suffix "-Column-Names") containing a key translating from the column names in this data file to column descriptions, and a text file (with the suffix "-Table-Notes") providing more information on how the data were collected.

We read in the data and column names files, then use the column names file to identify a subset of fields in the data file we will keep. Specifically, we choose to keep:

* Total Population
* Average Household Size
* Percent of population 25 and over with educated high school degree or greater
* Percent of population 25 and over educated with Bachelor's degree or greater
* Percent of civilian non-institutionalized population with a disability
* Average commute time (in minutes)
* Percent of civilian employed population 16 and over who are employed as private wage and salary workers
* Percent of civilian employed population 16 and over who are employed as government workers
* Percent of civilian employed population 16 and over who are employed as self-employed workers
* Percent of civilian employed population 16 and over who are employed as unpaid family workers
* Median household income
* Median individual per capita income
* Median value of owner-occupied household units
* Median rent

These fields all looked relevant to employment, and therefore commute time, while also being relatively interpretable. We also keep total counts that will allow us to compute weighted averages across counties when we analyze the data. However, for all fields, we exclude "margin of error" calculations for our analysis.

After filtering to these variables, we also remove the first row of the data, which is the same as the second column of the "Column-Names" file. Since this row consists of all character strings, we then convert the columns except for the county ID and name into numeric variables. Finally, we extract the substring of the county name that identifies the state in which the county is located.

We save the cleaned intermediate file in the data folder.

```{r}
#| label: load downloaded data

source(here("source", "01_load_data.R"))
```

# Data Analysis

Next, we summarize the data and run regressions to identify potential associations between these fields.

```{r}
#| label: data analysis

source(here("source", "02_analyze_data.R"))
```

## Data Summary

First, we summarize the data by state to get a broader view of the data. There are 39 states represented in the data, with each state contributing between 1 and 16 counties to the data set.

```{r}
#| label: state summary

# Print table of state summaries
kable(acs_state, caption = "State Summary", digits = 2,
      col.names = c("State", "Counties", "Mean Commute",
                    "Household Size",
                    "Pct High School Educated", "Pct Bachelor's Educated",
                    "Pct with Disability", 
                    "Pct. Privately Employed", "Pct. Government Employed", "Pct Self-Employed", "Pct. Family Employed", 
                    "Median Household Income", 
                    "Median Value of Owner-Occupied Units", "Median Rent"))
```

## Univariate Regressions

Next, we run univariate analyses on each of these variables to understand basic associations between the metrics and average commute time. We find that 7 of the 11 variables are significantly associated with average commute time at a significance level of $\alpha = 0.05$:

* Median Rent
* Median Value of Owner-Occupied Units
* Median Household Income
* Average Household Size
* Percent of the Population with High School Education or More
* Percent of the Population with a Disability
* Percent of the Population Self-Employed

On the other hand, the following 4 variables are not significantly associated with average commute time:

* Percent of the Population Privately Employed
* Percent of the Population with Bachelor's Degree or More
* Percent of the Population Employed by the Government
* Percent of the Population Employed by Unpaid Family Work

These results are shown in the table below in the order above, which is in increasing order of p-value.

```{r}
#| label: univariate regressions

kable(acs_reg_uni_tbl, digits = 3, 
      caption = "Univariate Regression Results",
      col.names = c("Variable", "Estimate", "P-Value", "Significance Level"))
```

## Multivariate Regressions

Then, we run a multivariate linear regression relating average commute time to the whole set of variables that were found significant in univariate regressions. In this multivariate regression, only 3 variables were found to be associated with average commute time given the remaining variables:

* Median Rent
* Percent of the Population with High School Education or More
* Median Value of Owner-Occupied Units

```{r}
#| label: display full reg

# Print clean table of regression results
tidy(acs_reg) %>% 
  arrange(term != "(Intercept)", p.value) %>% 
  mutate(p_signif = stars.pval(p.value),
         p.value = format.pval(p.value, digits = 3, eps = 0.001)) %>% 
  kable(caption = "Full Multivariate Regression Results",
        digits = 3)
```

In order to identify whether some of the non-significant variables in this model could be significant in a sparser model, we ran stepwise selection on this model using AIC. This procedure yielded a model including the three variables above, as well as the Percent of the Population Self-Employed.

These effects are shown in the table below.

```{r}
#| label: display sparse reg

# Print clean table of sparse regression results
tidy(acs_reg_sparse) %>% 
  arrange(term != "(Intercept)", p.value) %>% 
  mutate(p_signif = stars.pval(p.value),
         p.value = format.pval(p.value, digits = 3, eps = 0.001)) %>% 
  kable(caption = "Sparse Multivariate Regression Results",
        digits = c(0, 5, 3, 3, 0, 0))
```

# Discussion

To better understand these results, we visualize the data in two ways. 

```{r}
#| label: visualizations

source(here("source", "03_visualize_data.R"))
```

## Visualizations

First, to understand the univariate results, we create scatterplots to illustrate how each of the variables we include in our analysis are related to average commute time.

```{r}
#| label: scatterplots

acs_scatter
```

These scatterplots demonstrate that the variables that have the clearest relationships with commute time are average household size, median rent, median housing value of owner-occupied units, and median income. Other variables, such as those related to percent of the population employed by the government, private, or self-employed, show some clustering. 

Based on our analyses above, not all of these variables are strongly linearly associated with commute time after conditioning on all of the other variables. To understand how these variables may be related to each other, we also create a correlation plot, which we reorder based on hierarchical clustering.

```{r}
#| label: correlation-plot

corrplot(acs_corr, type = "lower", order = "hclust",
         tl.srt = 45, cl.pos = "b")
```

Based on this correlation plot, there are several variables that are correlated with each other. Median income, median housing value, median rent, and the prevalence of Bachelor's degree education or more are all strongly positively correlated. The prevalence of high school education level is less correlated with housing values, but is correlated with income and prevalence of Bachelor's degree education. On the other hand, the prevalence of disability in a county is strongly negatively correlated with housing values, income, prevalence of a Bachelor's degree, and average household size.

These correlations help explain why a variable like median income is not included in our final model, despite it having a strongly significant univariate association. Conditioned on housing values, it is likely not as strongly associated with average commute time.

The only variable that was not significant in the full multivariate regression but was significant in the sparse multivariate regression was the prevalence of self-employed workers. The correlation plot shows that this metric is moderately positively correlated with average household size, housing values, and prevalence of high school education. Therefore, it is likely that its correlation with average household size was masking its association with average commute time. A sparse model excludes average household size, so the negative association between self-employment and average commute time becomes more evident.

## Conclusion

This analysis shows that housing prices, educational attainment, and employment class are most strongly associated with commute time. Of these, only average rent is positively associated with average commute time in a county. This may point to the population density of each county, which could increase both rent price and traffic. Prevalence of high school education, median owner-occupied unit housing value, and prevalence of self-employment are all negatively associated with average commute time. These results are somewhat intuitive: education may allow residents to find employment closer to where they live, self-employment likely involves short commutes, and higher housing value may indicate that residents live closer to an economic hub. However, the univariate association between housing value and commute time was positive, so there could also be an interaction between median owner-occupied housing value and median rent.
